Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2018-08-07 paper/manuscript.tex
some updates to outline.

\begin{checklist}{Step 0 -- Know what you want to simulate }
\textbf{What are the first questions that need addressing before setting up a molecular dynamics simulation}\\
Extensive explanation for the checklist questions can be found in section~\ref{sec:step0}.
\begin{itemize}
\item Can I get the required accuracy with the simulation I want to carry out
\item
\item And finally
\end{itemize}
\end{checklist}

@@

\begin{Checklists*}[p!]
\begin{checklist}{Absolute simulations}
\textbf{What are the main things I need to consider for an absolute alchemical free energy calculation?}
Extensive explanation for the checklist questions can be found in section~\ref{sec:step2}.
\begin{itemize}
\item Topology
\item Restraints
\item Standard state handling
\end{itemize}
\end{checklist}

@@

\begin{checklist}{Analysis}
\textbf{This is all about analysis of the simulation}
Extensive explanation for the checklist questions can be found in section~\ref{sec:step4}.
\begin{itemize}
\item Are my simulations converged enough?
\item Am I using the right analysis techniques?
\end{itemize}
\end{checklist}

@@

\todo[inline, color={red!40}]{LNN: complete discussion about Boresch restraints}

\todo[inline, color={red!40}]{AR: would the explanation in the next paragraph be better suited for a more general section?}

@@

\paragraph{Structural definition of the bound state.}
\todo[inline, color={red!40}]{AR: This section presents concepts that are common to absolute and relative binding free energy calculations, but it is useful to have \textit{before} talking about restraints. Should we move the "common" section before the "differences" section?}

\paragraph{Determination of the reference state.}

Absolute free energies are typically reported with respect to a specific reference or standard state.
The reference state effectively determines the arbitrary point at which the binding free energy is 0, and it allow us to obtain a well-defined partial molar free energy for the binding reaction
\begin{equation*}
\ce{AB <=> A + B}
\end{equation*}
through the expression
\begin{equation} \label{eq:DGfromKAB}
\Delta G = -RT ~ \text{ln} \left( C^0 K_{AB} \right)  = -RT ~ \text{ln}\left( \frac{C^0 C_{AB}}{C_A C_B} \right) ,
\end{equation}
where $R$ is the gas, $T$ is the temperature, $C_X$ is the equilibrium concentration of the chemical species $X$ in the reaction solvent, and the reference state concentration $C^0$ converts the binding constant $K_{AB}$ into a dimensionless quantity expressed in reference concentration units.
It should be noted that ignoring the term $C^0$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used for $K_{AB}$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
Typically, it is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., A, B, and AB) in the reaction solvent has a concentration of $C^0$~=~1~M~=~1 molecule/1660 \r{A}$^3$ but do not interact with other molecules of A, B, or AB.

For solvation free energy calculations, this is typically straightforward to handle and treating it correctly simply means ensuring that the non-interacting solute is taken to the same (or equivalent) final reference state in both environments, e.g. that the transformation involves a 1M to 1M equivalent transfer free energy (where the non-interacting solute still occupies essentially the same volume as the solute in the interacting system).
So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state and if it does not, a simple entropic correction is needed.

However, for binding the situation is much more complex and requires special care.
Experimental absolute binding free energies are reported relative to a specific reference state -- a 1 M standard state -- which must also be used in calculations.
In practice, this has implications for how the calculations are done, as the reference concentration must enter the thermodynamic cycle employed.

Typically, to deal with both practical sampling issues and the standard state issue, restraints are employed in absolute binding free energy calculations to keep the ligand in a well defined volume as its interactions with the system are removed [ref Gilson 1997 BPJ].
This solves two problems.
First, if the ligand were not kept in a well-defined region, as its interactions were removed it might wander the system, perhaps quite slowly, and only inadequately sample the noninteracting or weakly interacting state -- yet adequate sampling of these states might be required for convergence.
So for practical purposes, the use of restraints can dramatically improve sampling as interactions are weakened and removed.

\todo[inline, color={red!40}]{AR: Is the next sentence true? I thought in this case the volume correction should just become the box volume unless the convergence is not reached, and this is covered by the previous paragraph. I thought the problem with not having the restraint is that you are sampling configurations that are not "bound", but here we enter the discussion of weak binders, as this problem is greater the weaker is the binding. Maybe we should present the problem of weak binders before this section as it is a problem with relative calculations as well.}

Second, if the ligand is not kept in a well-defined region then it is hard to determine how to link a computed binding free energy to the correct 1M standard state.
In contrast, with restraints, the free energy of releasing the restrained ligand to a 1M standard state can be computed analytically or numerically by solving the relevant integral [refs], allowing the standard state to enter the thermodynamic cycle [refs].

\todo[inline, color={red!40}]{AR: Given the audience, it might be good to discuss the connection to the experiments signal to define the bound state, viz. the experimental methodology must be sensitive to the configurations that are defined to be "bound" and its signal of the "unbound" configurations should be negligible, although this is not specific to absolute binding free energies. I'm now thinking we should discuss this in a previous section. Maybe in the "know what you want to simulate" step? In that paragraph, we should also mention that most modern calculations rely on kinetic traps, but that this doesn't work with weak binders or enhanced sampling techniques without proper restraints.}

@@

\section{Step 3 -- Overview of available analysis techniques}
\label{sec:step3}
\todo[inline, color={red!40}]{JDC, BA: uncertainty estimation section}
\begin{enumerate}
\item Detecting boundary between equilibrated and production regions (Chodera: \url{http://dx.doi.org/10.1021/acs.jctc.5b00784})
\item Decorrelating samples for analysis
\begin{itemize}
\item Subsample different lambdas based on correlation times
\item Ensure all simulations at least 50x correlation time
\end{itemize}
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\begin{itemize}
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More “usual suspects”
\end{itemize}
\item Estimators for free energies
\begin{itemize}
\item MBAR recommended if all energy differences are available
\item BAR just as good for highly optimized lambda values
\item TI should be roughly concordant, but quadrature error hard to quantify
\item Other variants useful in special circumstances (e.g. Z. Tan stochastic version)
\end{itemize}
\item Computing and reporting uncertainties on free energies
correlated bootstrap v. timeseries analysis
\begin{itemize}
\item Quantifying standard error in dG estimate
\item Varience in dG estimate using multiple methods (TI, DEXP, IEXP, BAR, MBAR, GDEL, etc.)
\item Agreement in dG estimate when repeating calculations with different parameters (random seeds, initial configurations, forcefield, etc.)
\item Calculating differences in free energy change as a function of time
\begin{itemize}
	\item Starting from beginning or end of simulation
	\item Significance of differences in midpoint estimate (High middle error: high uncertainty)
\end{itemize}
\item Ensemble method to combine uncertainties into interpretable weighted metric
\begin{itemize}
	\item Simple normalized metric to determine confidence in calculation
	\item Easily interpreted by chemist/biologist when prioritizing new chemistries
\end{itemize}
\end{itemize}
\item Other considerations for many transformations
Cycle closure error

>>>>

\begin{checklist}{Step 0 -- Know what you want to simulate }
\textbf{What are the first questions that need addressing before setting up a molecular dynamics simulation}\\
Extensive explanation for the checklist questions can be found in section~\ref{sec:step0}.
\begin{itemize}
\item Can I get the required accuracy with the simulation I want to carry out
\item
\item And finally
\item choice of equilibrium or nonequilibrium methods (probably prefer equilibrium, because easier to check things; nonequilibrium requires checking to make sure the distribution is converged, but you aren't guaranteed you don't miss things.
\end{itemize}
\end{checklist}

@@

\begin{Checklists*}[p!]
\begin{checklist}{Absolute simulations}
\textbf{What are the main things I need to consider for an absolute alchemical free energy calculation?}
Extensive explanation for the checklist questions can be found in section~\ref{sec:step2}.
\begin{itemize}
\item Topology
\item Restraints
\item Standard state handling
\end{itemize}
\item Hamiltonian Replica exchange can always help, if you can run the simulations in parallel, as the simulation will be able to sample more in the uncoupled state.
\end{checklist}

@@

\begin{checklist}{Analysis}
\textbf{This is all about analysis of the simulation}
Extensive explanation for the checklist questions can be found in section~\ref{sec:step4}.
\item Pick an appropriate method; usually MBAR, BAR, or TI.  Better yet, use multiple types of analysis.
\item Make sure the data you collect are uncorrelated samples.
\item Visualize the molecule to make sure it is behaving as you expect.
\end{checklist}

@@

\todo[inline, color={red!40}]{AR, LNN: complete absolute free energy simulation protocol}

@@

Absolute free energy calculations are typically done with respect to a specific reference state or reference concentration.
For solvation free energy calculations this is typically straightforward to handle and treating it correctly simply means ensuring that the non-interacting solute is taken to the same (or equivalent) final reference state in both environments, e.g. that the transformation involves a 1 M to 1M equivalent transfer free energy (where the non-interacting solute still occupies essentially the same volume as the solute in the interacting system).
So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state and if it does not, a simple entropic correction is needed.

However, for binding the situation is much more complex and requires special care.
Experimental absolute binding free energies are reported relative to a specific reference state -- a 1 M standard state -- which must also be used in calculations.
In practice this has implications for how the calculations are done, as the reference concentration must enter the thermodynamic cycle employed.

Typically, to deal with both practical sampling issues and the standard state issue, restraints are employed in absolute binding free energy calculations to keep the ligand in a well defined volume as its interactions with the system are removed [ref Gilson 1997 BPJ].
This solves two problems.
First, if the ligand were not kept in a well-defined region, as its interactions were removed it might wander the system, perhaps quite slowly, and only inadequately sample the noninteracting or weakly interacting state -- yet adequate sampling of these states might be required for convergence.
So for practical purposes, the use of restraints can dramatically improve sampling as interactions are weakened and removed.
Second, if the ligand is not kept in a well-defined region then it is hard to determine how to link a computed binding free energy to the correct 1M standard state.
In contrast, with restraints, the free energy of releasing the restrained ligand to a 1M standard state can be computed analytically or numerically by solving the relevant integral [refs], allowing the standard state to enter the thermodynamic cycle [refs].

@@

\section{Step 3 -- Overview of available analysis techniques}
\label{sec:step3}
\todo[inline, color={red!40}]{JDC, BA: uncertainty estimation section}
\begin{enumerate}
\item Detecting boundary between equilibrated and production regions (Chodera: \url{http://dx.doi.org/10.1021/acs.jctc.5b00784})
\item Decorrelating samples for analysis
\begin{itemize}
\item Subsample different lambdas based on correlation times
\item Ensure all simulations at least 50x correlation time %MRS: I've heard 40x from Swope. Rationale has to do with the number of uncorrelated samples for the estimate of the uncertainty to be reliable.
\end{itemize}
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\begin{itemize}
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item If possible, look at multiple estimators.  Significant discrepancies between different free energy estimators (TI, BAR, MBAR) indicate there is a problem..
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation (best practice would be to look at 
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More ``usual suspects''
\end{itemize}
\item Estimators for free energies
\begin{itemize}
\item MBAR is gold standard, but requires energies from each sample at all of the other states.  BAR is almost as good if the states are separated so as to minimize the number of states (i.e. there is good overlap from $i \rightarrow i+1$ and $i \rightarrow i-1$, but not to $i\pm 2$.
\item Using bootstrap (or block bootstrap) for errors is more reliable than analytical error estimates. BAR uncertainties are underestimate.
\item TI should be roughly in agreement, but quadrature error hard to quantify; should generally be used only if the path is known to be good for that size of molecule.
\item Other variants useful in special circumstances (e.g. Z. Tan stochastic version)
\item expontential averaging is very problematic.  If using intermediates (which will almost always be the case) should avoid.
\end{itemize}
\item Computing and reporting uncertainties on free energies
\item correlated bootstrap versus timeseries subsampling.
\begin{itemize}
\item Quantifying standard error in dG estimate
\item Variance in dG estimate using multiple methods (TI, DEXP, IEXP, BAR, MBAR, GDEL, etc.)
\item Agreement in dG estimate when repeating calculations with different parameters (random seeds, initial configurations, forcefield, etc.)
\item looking at the overlap matrix and making sure that there is decent overlap between all states.
\item Calculating differences in free energy change as a function of time
\begin{itemize}
	\item Starting from beginning or end of simulation
	\item Significance of differences in midpoint estimate (High middle error: high uncertainty)
\end{itemize}
\item Ensemble method to combine uncertainties into interpretable weighted metric
\begin{itemize}
	\item Simple normalized metric to determine confidence in calculation
	\item Easily interpreted by chemist/biologist when prioritizing new chemistries
\end{itemize}
\end{itemize}
\item Other considerations for many transformations
\begin{itemize}
Compute cycle closure error
\end{itemize}
